---
title: "Harnessing Data from North Carolina's Jails to Inform Effective Policies — California Jail Profile Survey"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
```

```{r data}
county_m <- read.csv("california_jail_county_monthly_1995_2020.csv")
county_q <- read.csv("california_jail_county_quarterly_1995_2020.csv")
facility <- read.csv("california_jail_facility_monthly_1995_2020.csv")
```

These datasets are scraped from the California Board of State and Community Corrections' Jail Profile Survey by Jacob Kaplan from the University of Pennsylvania. Kaplan made the following changes to the data: adding variables for FIPS county and state codes and United States census county names, changing column names, and changing values from "Does Not Apply" and "Unavailable" to NA. 

# County-Level Monthly Data

The county-level monthly data contains 17801 observations of 48 variables, with data collected from 58 counties from October 1995 to March 2020. Here is the first observation.

```{r county_monthly_dataset}
head(county_m, n = 1)
```

## Proportion of Missingness

22 of the 48 variables are missing in at least 10% of the observations. Here are the 10 most missing variables.

```{r county_monthly_proportion_missingness}
enframe(colMeans(is.na(county_m))) %>% 
  arrange(desc(value)) %>% 
  top_n(10)
```

## Class of Variables

All variables seem to have the correct data type, except date and day_of_highest_count, which represents the day of the month when the county had the highest count of inmates. We change those variables to Date variables.

```{r county_monthly_class}
enframe(sapply(county_m, class))
county_m$date <- as.Date(county_m$date, format = "%Y-%m-%d")
county_m$day_of_highest_count <- as.Date(county_m$day_of_highest_count, format = "%Y-%m-%d")
```

## Analysis of Missingness
```{r county_monthly_missingness_analysis, results = FALSE}
# variables with missingness concentrated by county
table(county_m$census_county_name, is.na(county_m$avg_felony_inmate_unsentenced))
table(county_m$census_county_name, is.na(county_m$avg_misdemean_inmate_unsentenced))
table(county_m$census_county_name, is.na(county_m$avg_own_inmate_housed_elsewhere))
table(county_m$census_county_name, is.na(county_m$avg_fed_inmate_housed_contract))
table(county_m$census_county_name, is.na(county_m$avg_state_inmate_housed_contract))
table(county_m$census_county_name, is.na(county_m$avg_local_inmate_housed_contract))

# number & proportion of missingness per county if missing more than 0.1
county_m %>% 
  group_by(census_county_name) %>% 
  count(is.na(avg_felony_inmate_unsentenced)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(avg_felony_inmate_unsentenced)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_m %>% 
  group_by(census_county_name) %>% 
  count(is.na(avg_misdemean_inmate_unsentenced)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(avg_misdemean_inmate_unsentenced)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_m %>% 
  group_by(census_county_name) %>% 
  count(is.na(avg_own_inmate_housed_elsewhere)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(avg_own_inmate_housed_elsewhere)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_m %>% 
  group_by(census_county_name) %>% 
  count(is.na(avg_fed_inmate_housed_contract)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(avg_fed_inmate_housed_contract)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_m %>% 
  group_by(census_county_name) %>% 
  count(is.na(avg_state_inmate_housed_contract)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(avg_state_inmate_housed_contract)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_m %>% 
  group_by(census_county_name) %>% 
  count(is.na(avg_local_inmate_housed_contract)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(avg_local_inmate_housed_contract)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

# before 2002 missingness
table(county_m$year, is.na(county_m$adp_of_maximum_security_inmates))
table(county_m$year, is.na(county_m$adp_of_medium_security_inmates))
table(county_m$year, is.na(county_m$adp_of_minimum_security_inmates))
table(county_m$year, is.na(county_m$avg_inmates_get_medical_bed))
table(county_m$year, is.na(county_m$avg_inmates_get_mental_heath_bed))
table(county_m$year, is.na(county_m$avg_inmate_need_reg_med_attent))
table(county_m$year, is.na(county_m$avg_inmates_need_reg_ment_health))

# after 2002 missingness
table(county_m$year, is.na(county_m$num_inmate_get_psych_meds))
table(county_m$year, is.na(county_m$num_inmate_get_mental_heath_bed))
table(county_m$year, is.na(county_m$avg_inmate_get_sick_bed))
table(county_m$year, is.na(county_m$num_inmates_seen_sick_call))
table(county_m$year, is.na(county_m$num_doctor_occurrences))
table(county_m$year, is.na(county_m$num_dental_encounters))
table(county_m$year, is.na(county_m$num_offsite_medical_appointment))

county_m %>% 
  filter(year > 2001) %>% 
  group_by(census_county_name) %>%  
  count(is.na(num_inmate_get_psych_meds)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(num_inmate_get_psych_meds)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_m %>% 
  filter(year > 2001) %>% 
  group_by(census_county_name) %>%  
  count(is.na(num_inmate_get_mental_heath_bed)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(num_inmate_get_mental_heath_bed)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_m %>% 
  filter(year > 2001) %>% 
  group_by(census_county_name) %>%  
  count(is.na(avg_inmate_get_sick_bed)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(avg_inmate_get_sick_bed)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_m %>% 
  filter(year > 2001) %>% 
  group_by(census_county_name) %>%  
  count(is.na(num_inmates_seen_sick_call)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(num_inmates_seen_sick_call)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_m %>% 
  filter(year > 2001) %>% 
  group_by(census_county_name) %>%  
  count(is.na(num_doctor_occurrences)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(num_doctor_occurrences)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_m %>% 
  filter(year > 2001) %>% 
  group_by(census_county_name) %>%  
  count(is.na(num_dental_encounters)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(num_dental_encounters)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_m %>% 
  filter(year > 2001) %>% 
  group_by(census_county_name) %>%  
  count(is.na(num_offsite_medical_appointment)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(num_offsite_medical_appointment)` & proportion > 0.1) %>% 
  arrange(desc(proportion))
```

The following counties are missing at least 10% of data for average number of misdemeanor and felony inmates unsentenced: Del Norte, Imperial, Inyo, Lassen, Monterey, Napa, San Benito, San Diego, Santa Barbara, Santa Clara, Santa Cruz, Shasta, Tehama, Trinity, Tulare, Tuolumne, and Ventura. Tehama and Santa Barbara are missing, respectively, at least 96% and 42% of data for those variables. Although missingness statewide is slightly concentrated before 2001, it is more concentrated by county.

The following variables were not counted after 2002: average daily population of maximum, medium, and minimum security inmates, average number of inmates assigned to medical and mental health beds, and average number of inmates needing regular medical and mental health attention.

The following variables were not counted before 2002: number of inmates receiving psychiatric mediation on the last day of the month, number of inmates assigned to medical and mental health beds on the last day of the month, number of inmates seen at sick call, number of doctor occurences, number of dental encounters, number of offsite medical appointments.

# County-Level Quarterly Data

The county-level quarterly data contains 5905 observations of 18 variables, with data collected from 58 counties from October 1995 to March 2020. Here is the first observation.

```{r county_quarterly_dataset}
head(county_q, n = 1)
```

## Proportion of Missingness

10 of the 18 variables are missing in at least 10% of the observations. Here are the 10 most missing variables.

```{r county_quarterly_proportion_missingness}
enframe(colMeans(is.na(county_q))) %>% 
  arrange(desc(value)) %>% 
  top_n(10)
```

## Class of Variables

All variables seem to have the correct data type, except quarter, which we change to a Factor variable.

```{r county_quarterly_class}
sapply(county_q, class)
county_q$quarter <- as.factor(county_q$quarter)
```

## Analysis of Missingness
```{r county_quarterly_missingness_analysis, results = FALSE}
table(county_q$year, is.na(county_q$num_inmate_are_2nd_striker))
table(county_q$year, is.na(county_q$num_inmate_are_3rd_striker))
table(county_q$year, is.na(county_q$num_unserved_misdemean_warrants))
table(county_q$year, is.na(county_q$num_unserved_felony_warrants))
table(county_q$year, is.na(county_q$percent_inmates_illegal_alien))

table(county_q$year, is.na(county_q$avg_length_stay_pretrial_release))
table(county_q$year, is.na(county_q$avg_length_stay_sentnced_release))
table(county_q$year, is.na(county_q$avg_length_stay_all_releases))
table(county_q$year, is.na(county_q$money_spent_psych_med_last_qtr))
table(county_q$year, is.na(county_q$money_spent_medication_last_qtr))

# before 2001 missingness
county_q %>% 
  filter(year < 2002) %>% 
  group_by(census_county_name) %>% 
  count(is.na(num_inmate_are_2nd_striker)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(num_inmate_are_2nd_striker)` & proportion > 0.1)

county_q %>% 
  filter(year < 2002) %>% 
  group_by(census_county_name) %>% 
  count(is.na(num_inmate_are_3rd_striker)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(num_inmate_are_3rd_striker)` & proportion > 0.1)

county_q %>% 
  filter(year < 2002) %>% 
  group_by(census_county_name) %>% 
  count(is.na(num_unserved_misdemean_warrants)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(num_unserved_misdemean_warrants)` & proportion > 0.1)

county_q %>% 
  filter(year < 2002) %>% 
  group_by(census_county_name) %>% 
  count(is.na(num_unserved_felony_warrants)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(num_unserved_felony_warrants)` & proportion > 0.1)

county_q %>% 
  filter(year < 2002) %>% 
  group_by(census_county_name) %>% 
  count(is.na(percent_inmates_illegal_alien)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(percent_inmates_illegal_alien)` & proportion > 0.1)

# after 2001 missingness
county_q %>% 
  filter(year > 2001) %>% 
  group_by(census_county_name) %>%  
  count(is.na(avg_length_stay_pretrial_release)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(avg_length_stay_pretrial_release)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_q %>% 
  filter(year > 2001) %>% 
  group_by(census_county_name) %>% 
  count(is.na(avg_length_stay_sentnced_release)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(avg_length_stay_sentnced_release)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_q %>% 
  filter(year > 2001) %>% 
  group_by(census_county_name) %>% 
  count(is.na(avg_length_stay_all_releases)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(avg_length_stay_all_releases)` & proportion > 0.1) %>% 
  arrange(desc(proportion))

county_q %>% 
  filter(year > 2001) %>% 
  group_by(census_county_name, year) %>% 
  count(year, is.na(money_spent_psych_med_last_qtr)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(money_spent_psych_med_last_qtr)` & proportion > 0.1) %>% 
  arrange(desc(proportion)) %>% 
  group_by(census_county_name) %>% 
  count(census_county_name)

county_q %>% 
  filter(year > 2001) %>% 
  group_by(census_county_name, year) %>% 
  count(is.na(money_spent_medication_last_qtr)) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(`is.na(money_spent_medication_last_qtr)` & proportion > 0.1) %>% 
  arrange(desc(proportion)) %>% 
  group_by(census_county_name) %>% 
  count(census_county_name)
```

The following variables were not counted after 2001: number of inmates who are 2nd strikers, number of inmates who are 3rd strikers, number of unserved misdemeanor warrants in county, number of unserved felony warrants in county, and percent of inmates who are undocumented immigrants. I subset the data, filtering for observations before 2001 and grouping by county. 

The following variables were not counted before 2001: average length of stay for pre-trial release, average length of stay for sentenced release, average length of stay for all releases, money spent on psychiatric medicine during the previous quarter, and money spent on medication during the previous quarter. I subset the data, filtering for observations after 2001 and grouping by county. In general, I try to identify randomness or concentration of missingness greater than 0.1.

For average length of stay for all releases, the following counties are missing data for every year: Contra Costa, Lake, San Luis Obispo, Santa Cruz, Sutter, Tuolumme, Yolo, and Yuba. For average length of stay for sentenced release, the same preceding counties, as well as Calaveras and Nevada, are missing data for every year. For average length of stay for pre-trial release, the same preceding counties, as well as Ventura, are missing data for every year.

# Facility/Jail-Level Monthly Data

The facility/jail-level monthly data contains 36774 observations of 14 variables, with data collected from 58 counties from October 1995 to March 2020. Here is the first observation.

```{r facility_dataset}
head(facility, n = 1)
```

## Proportion of Missingness

No variables are missing in at least 10% of the observations. Here are the 10 most missing variables.

```{r facility_proportion_missingness}
enframe(colMeans(is.na(facility))) %>% 
  arrange(desc(value)) %>% 
  top_n(10)
```

## Class of Variables

All variables seem to have the correct data type, except date, which we change to a Date variable.

```{r facility_class}
sapply(facility, class)
facility$date <- as.Date(facility$date, format = "%Y-%m-%d")
```

## Analysis of Missingness

```{r facility_missingness_analysis, results = FALSE}
# randomness & concentration of missingness for each variable
table(facility$year, is.na(facility$avg_daily_pop_total_facility))
table(facility$year, is.na(facility$avg_daily_pop_sentenced_male))
table(facility$year, is.na(facility$avg_daily_pop_unsentenced_male))
table(facility$year, is.na(facility$avg_daily_pop_sentenced_female))
table(facility$year, is.na(facility$avg_daily_pop_unsentenced_female))

facility %>% 
  filter(year == 2018 & (is.na(avg_daily_pop_total_facility) | 
           is.na(avg_daily_pop_sentenced_male) | is.na(avg_daily_pop_unsentenced_male) |
           is.na(avg_daily_pop_sentenced_female) | is.na(avg_daily_pop_unsentenced_female)))

facility %>% 
  filter(year < 2001 & is.na(avg_daily_pop_sentenced_male)) %>% 
  count(census_county_name, facility)

facility %>% 
  filter(year < 2001 & is.na(avg_daily_pop_unsentenced_male)) %>% 
  count(census_county_name, facility)

facility %>% 
  filter(year < 2001 & is.na(avg_daily_pop_sentenced_female)) %>% 
  count(census_county_name, facility)

facility %>% 
  filter(year < 2001 & is.na(avg_daily_pop_unsentenced_female)) %>% 
  count(census_county_name, facility)
```

The following counties are missing data for average daily populations from 2018: Trinity (February through June), Tehama (July), Siskiyou (September), and Sutter (September). All of the counties are in northern California, and their jail data collection efforts may have been affected by the 2018 California wildfires. 

Additional counties are missing data for average daily populations from 2000 or before, but some missingness is not random. For example, Santa Barbara County's Female Honor Farm is missing over 20 counts for sentenced and unsentence males, which is due to it being a facility for women.